---
title: 'Data Analysis Assignment #1 (50 points total)'
author: "roy, deblina"
output:
  html_document: default
---

```{r setup, include = FALSE}
# DO NOT ADD OR REVISE CODE HERE
knitr::opts_chunk$set(echo = FALSE, eval = TRUE)

```


Submit both the .Rmd and .html files for grading. You may remove the instructions and example problem above, but do not remove the YAML metadata block or the first, "setup" code chunk.  Address the steps that appear below and answer all the questions. Be sure to address each question with code and comments as needed.  You may use either base R functions or ggplot2 for the visualizations.

-----

The following code chunk will:

(a) load the "ggplot2", "gridExtra" and "knitr" packages, assuming each has been installed on your machine,
(b) read-in the abalones dataset, defining a new data frame, "mydata,"
(c) return the structure of that data frame, and
(d) calculate new variables, VOLUME and RATIO.

Do not include package installation code in this document. Packages should be installed via the Console or 'Packages' tab. You will also need to download the abalones.csv from the course site to a known location on your machine. Unless a *file.path()* is specified, R will look to directory where this .Rmd is stored when knitting.

```{r analysis_setup1, message = FALSE, warning = FALSE}

# a) Load the ggplot2 and gridExtra packages.

library(ggplot2)
library(gridExtra)
library(knitr)

# b) Use read.csv() to read the abalones.csv into R, assigning the data frame to "mydata."

mydata <- read.csv("abalones.csv", sep = ",", stringsAsFactors = TRUE)


# c) Use the str() function to verify the structure of "mydata." You should have 1036 observations
# of eight variables.

str(mydata)

# d) Define two new variables, VOLUME and RATIO. Use the following statements to define VOLUME and
# RATIO as variables appended to the data frame "mydata."

mydata$VOLUME <- mydata$LENGTH * mydata$DIAM * mydata$HEIGHT
mydata$RATIO <- mydata$SHUCK / mydata$VOLUME

```

-----

### Test Items starts from here - There are 6 sections - Total 50 points

***##### Section 1: (6 points) Summarizing the data.***

(1)(a) (1 point) Use *summary()* to obtain and present descriptive statistics from mydata.  Use *table()* to present a frequency table using CLASS and RINGS. There should be 115 cells in the table you present.  

```{r Part_1a}
summary(mydata)
table(mydata$CLASS, mydata$RINGS)

```

**Question (1 point):  Briefly discuss the variable types and distributional implications such as potential skewness and outliers.**

***Answer: ( 
The dataset contains both categorical and numerical variables. 
SEX and CLASS are categorical variables—SEX is nominal and balanced across categories (F, M, I), while CLASS is ordinal and Distribution is concentrated in middle age groups, not strongly skewed. 
Continuous variables include LENGTH, DIAM, and HEIGHT, all showing nearly symmetric distributions with close mean and median values and no major skewness or outliers. 
Variables like WHOLE, SHUCK, and VOLUME are continuous and show signs of positive skew and possible upper-end outliers due to large value ranges. 
RINGS is a discrete numeric variable with a slight positive skew indicating more young abalones than older ones. 
RATIO, though also continuous, has a narrow range and slight positive skew but no extreme outliers. 
Overall, while most variables are symmetric, skewness and outliers appear in weight-related measures (WHOLE, SHUCK, VOLUME) and age (RINGS).

)***

(1)(b) (1 point) Generate a table of counts using SEX and CLASS. Add margins to this table (Hint: There should be 15 cells in this table plus the marginal totals. Apply *table()* first, then pass the table object to *addmargins()* (Kabacoff Section 7.2 pages 144-147)).  Lastly, present a barplot of these data; ignoring the marginal totals.  

```{r Part_1b}
tab_sex_class <- table(mydata$SEX, mydata$CLASS)
addmargins(tab_sex_class)

barplot(tab_sex_class,
        beside = TRUE, 
        col = c("blue", "pink", "green"),
        legend = rownames(tab_sex_class),
        main = "counts by SEX and CLASS", 
        xlab = "CLASS",
        ylab = "Count")
      
```

**Essay Question (2 points):  Discuss the sex distribution of abalones.  What stands out about the distribution of abalones by CLASS?**

***Answer: ( 
The sex distribution of abalones across CLASS categories reveals important patterns related to growth, development, and maturation:

 1.	Infants Dominate Early Classes:
In Class A1, the vast majority of abalones are labeled as Infants (I). This strongly suggests that A1 corresponds to the earliest developmental stage. Class A2 still has a large number of Infants, but Males (M) begin to appear in higher numbers, indicating the start of sexual differentiation.
	2.	Transition to Adulthood in Mid Classes:
In Class A3, the distribution shifts significantly - Males and Females (F) surpass Infants in count. This class appears to be a transitional stage, where abalones mature from the Infant stage to identifiable adult sexes. Notably, Males outnumber Females, suggesting either a biological tendency or a sampling effect.
	3.	Adult Stability in Higher Classes:
Classes A4 and A5 show minimal Infant presence. These classes are predominantly composed of Males and Females, with Males slightly more prevalent. This points to a more stable sex distribution among mature abalones.
	4.	Sex Imbalance Across Classes:
While overall the sexes are somewhat balanced, there is a consistent trend of Males being more numerous than Females in older classes (A3 to A5). This could reflect faster growth, better survival rates, or differences in classification accuracy between sexes.

What stands out is that the dataset clearly reflects a biological progression:
	- Infants are concentrated in early classes,
	- Adults emerge by Class A3,
	- And by Classes A4 and A5, the population is composed almost entirely of Males and Females, with Males becoming slightly dominant.
)***

(1)(c) (1 point) Select a simple random sample of 200 observations from "mydata" and identify this sample as "work." Use *set.seed(123)* prior to drawing this sample. Do not change the number 123. Note that *sample()* "takes a sample of the specified size from the elements of x." We cannot sample directly from "mydata." Instead, we need to sample from the integers, 1 to 1036, representing the rows of "mydata." Then, select those rows from the data frame (Kabacoff Section 4.10.5 page 87). 

Using "work", construct a scatterplot matrix of variables 2-6 with *plot(work[, 2:6])* (these are the continuous variables excluding VOLUME and RATIO). The sample "work" will not be used in the remainder of the assignment. 

```{r Part_1c}
set.seed(123)
sample_rows <- sample(1:1036, size = 200)
work <- mydata[sample_rows, ]
plot(work[, 2:6])
```

-----

***##### Section 2: (5 points) Summarizing the data using graphics.***

(2)(a) (1 point) Use "mydata" to plot WHOLE versus VOLUME. Color code data points by CLASS.

```{r Part_2a}
# Plot WHOLE vs VOLUME with points color-coded by CLASS
library(ggplot2)

ggplot(mydata, aes(x = WHOLE, y = VOLUME, color = CLASS)) +
  geom_point() +  # Add points
  labs(title = "WHOLE vs VOLUME Colored by CLASS", x = "WHOLE", y = "VOLUME") +  # Add labels
  theme_minimal()  # Minimal theme for clean look

```


(2)(b) (2 points) Use "mydata" to plot SHUCK versus WHOLE with WHOLE on the horizontal axis. Color code data points by CLASS. As an aid to interpretation, determine the maximum value of the ratio of SHUCK to WHOLE.  Add to the chart a straight line with zero intercept using this maximum value as the slope of the line. If you are using the 'base R' *plot()* function, you may use *abline()* to add this line to the plot. Use *help(abline)* in R to determine the coding for the slope and intercept arguments in the functions. If you are using ggplot2 for visualizations, *geom_abline()* should be used.

```{r Part_2b}
# Load ggplot2 (though it's been loaded before but added it here again)
library(ggplot2)

# Ensure VOLUME and RATIO are defined if not already
mydata$VOLUME <- mydata$LENGTH * mydata$DIAM * mydata$HEIGHT
mydata$RATIO <- mydata$SHUCK / mydata$VOLUME

# Calculate max ratio of SHUCK to WHOLE
max_ratio <- max(mydata$SHUCK / mydata$WHOLE, na.rm = TRUE)

# Create the plot
ggplot(mydata, aes(x = WHOLE, y = SHUCK, color = CLASS)) +
  geom_point() +
  geom_abline(intercept = 0, slope = max_ratio, color = "black", linetype = "dashed") +
  labs(title = "SHUCK vs WHOLE Colored by CLASS",
       x = "Whole Weight",
       y = "Shucked Weight") +
  theme_minimal()

```

**Essay Question (2 points):  How does the variability in this plot differ from the plot in (a)?  Compare the two displays.  Keep in mind that SHUCK is a part of WHOLE.  Consider the location of the different age classes.**

***Answer: (
Plot 2(a): WHOLE vs VOLUME
		- Structure of the Relationship:
	 WHOLE and VOLUME have a strong positive correlation, but the relationship is less tightly linear.
	 The points show wider vertical spread (greater variability in VOLUME for a given WHOLE weight).
	  - Interpretation of Variability:
	Since VOLUME is derived from physical dimensions (LENGTH × DIAM × HEIGHT), while WHOLE is a weight, these two measures are related but not perfectly proportionate.
	Shape variation in abalones (e.g., same weight but different shapes) introduces natural variability in VOLUME.
	Different age classes (CLASS A1 to A5) are somewhat scattered across the WHOLE-VOLUME relationship, with older classes (A4, A5) tending to cluster at higher values.

Plot 2(b): SHUCK vs WHOLE
	 - Structure of the Relationship:
	This plot shows a tighter, bounded linear relationship, because SHUCK is a component of WHOLE.
	The maximum possible SHUCK cannot exceed WHOLE, which naturally restricts the range of the SHUCK values. The dashed line reflects the max SHUCK-to-WHOLE ratio as a constraint.
	 - Interpretation of Variability:
	Much less scatter compared to plot 2(a); data points are tightly concentrated below the dashed line.
	Indicates that SHUCK is highly dependent on WHOLE, but with some variation based on internal composition.
	Age classes (CLASS) are more clearly separated vertically: as abalones age, both SHUCK and WHOLE increase, but within a more predictable range than in the VOLUME case.

)***
-----

***### Section 3: (8 points) Getting insights about the data using graphs.***

(3)(a) (2 points) Use "mydata" to create a multi-figured plot with histograms, boxplots and Q-Q plots of RATIO differentiated by sex. This can be done using *par(mfrow = c(3,3))* and base R or *grid.arrange()* and ggplot2. The first row would show the histograms, the second row the boxplots and the third row the Q-Q plots. Be sure these displays are legible.  

```{r Part_3a}

# Ensure RATIO and SEX are defined properly
mydata$SEX <- as.factor(mydata$SEX)

# Define the levels of SEX
sexes <- levels(mydata$SEX)

# Set up 3 rows × 3 columns plotting layout
par(mfrow = c(3, 3), mar = c(4, 4, 2, 1))

# Row 1: Histograms of RATIO by SEX
for (s in sexes) {
  hist(mydata$RATIO[mydata$SEX == s],
       main = paste("Histogram of RATIO -", s),
       xlab = "RATIO", col = "skyblue", border = "white")
}

# Row 2: Boxplots of RATIO by SEX
for (s in sexes) {
  boxplot(mydata$RATIO[mydata$SEX == s],
          main = paste("Boxplot of RATIO -", s),
          ylab = "RATIO", col = "lightgreen")
}

# Row 3: Q-Q plots of RATIO by SEX
for (s in sexes) {
  qqnorm(mydata$RATIO[mydata$SEX == s],
         main = paste("Q-Q Plot of RATIO -", s))
  qqline(mydata$RATIO[mydata$SEX == s], col = "red")
}
```

**Essay Question (2 points): Compare the displays.  How do the distributions compare to normality? Take into account the criteria discussed in the sync sessions to evaluate non-normality.**

***Answer: (
 Histogram Observations:
	-	All three histograms (F, I, M) show slight right-skewness, meaning the distribution tails off to the right with more data concentrated toward the left.
	-	The distributions are unimodal but not perfectly symmetric, especially for Infants, where a small bump at the higher end suggests potential bimodality.

 Boxplot Observations:
	- Each sex group shows visible outliers (above the upper whisker), particularly among Infants and Females.
	- The median lines are not centered in the boxes, further indicating skewness.
	- Infants exhibit more spread and a wider interquartile range, suggesting more variability in the RATIO.

 Q-Q Plot Observations:
	- For all three sexes, the Q-Q plots show deviations from the straight red line, especially in the tails:
	-	Heavy tails are present (points deviate at the ends), meaning the data have more extreme values than a normal distribution would expect.
	-	S-shaped patterns confirm departures from normality—indicating skewness or kurtosis.
	-	Infants have the most pronounced departure from normality.

 Overall,
	- The RATIO variable does not follow a normal distribution for any of the sex categories.
	- There is moderate right-skewness, especially for Infants, and outliers that further distort normality assumptions.
	- Based on criteria discussed in sync session (e.g. symmetry, absence of heavy tails), none of the groups can be assumed to be normally distributed. Therefore, it is recommended to consider transformations or apply non-parametric methods if the analysis requires the assumption of normality.
)***

(3)(b) (2 points) The boxplots in (3)(a) indicate that there are outlying RATIOs for each sex. *boxplot.stats()* can be used to identify outlying values of a vector. Present the abalones with these outlying RATIO values along with their associated variables in "mydata". Display the observations by passing a data frame to the kable() function. Basically, we want to output those rows of "mydata" with an outlying RATIO, but we want to determine outliers looking separately at infants, females and males.

```{r Part_3b}
library(knitr)

# Store outlier rows in a list (faster than rbind in loop)
outlier_rows_list <- list()

for (s in levels(mydata$SEX)) {
  subset_sex <- subset(mydata, SEX == s)
  outlier_vals <- boxplot.stats(subset_sex$RATIO)$out
  outlier_rows_list[[s]] <- subset_sex[subset_sex$RATIO %in% outlier_vals, ]
}

# Combine into one data frame
outliers_data <- do.call(rbind, outlier_rows_list)

# Display as table
kable(outliers_data, caption = "Outlying RATIO values by Sex")

```

**Essay Question (2 points):  What are your observations regarding the results in (3)(b)?**

***Answer: (
The analysis in (3)(b) identifies outliers in the RATIO variable (defined as SHUCK weight divided by VOLUME) separately for each sex (F, M, I). These outliers are extreme values in their respective groups.
 Females (F):
	- Most outliers in females have very large SHUCK weights relative to their volume, pushing their RATIO values higher.
	- One female (F.586) stands out as a low-end outlier with a very small RATIO (~0.067), despite having a large VOLUME and SHUCK weight. This could be due to mismeasurement or a biologically atypical case.
Infants (I):
	-	Outliers here tend to have very small body sizes (e.g., LENGTH < 5 cm) but relatively higher SHUCK weights, making their RATIO higher.
	- Since infants are not fully developed, small variations in SHUCK weight can greatly impact the RATIO, explaining their sensitivity to outlier detection.
Males (M):
	-	Male outliers have moderately large volumes and high SHUCK weights, indicating better-than-average yield.
	-	Their RATIO values are more tightly clustered, suggesting less variability in body efficiency among males compared to infants.
)***


***### Section 4: (8 points) Getting insights about possible predictors.***

(4)(a) (3 points) With "mydata," display side-by-side boxplots for VOLUME and WHOLE, each differentiated by CLASS There should be five boxes for VOLUME and five for WHOLE. Also, display side-by-side scatterplots:  VOLUME and WHOLE versus RINGS. Present these four figures in one graphic:  the boxplots in one row and the scatterplots in a second row. Base R or ggplot2 may be used.


```{r Part_4a}
# Set up a 2x2 plot layout
par(mfrow = c(2, 2),   # 2 rows, 2 columns
    mar = c(4, 4, 2, 1))  # Margins: bottom, left, top, right

# Boxplot 1: VOLUME by CLASS
boxplot(VOLUME ~ CLASS, data = mydata,
        main = "Boxplot of VOLUME by CLASS",
        xlab = "CLASS", ylab = "VOLUME",
        col = "lightblue")

# Boxplot 2: WHOLE by CLASS
boxplot(WHOLE ~ CLASS, data = mydata,
        main = "Boxplot of WHOLE by CLASS",
        xlab = "CLASS", ylab = "WHOLE",
        col = "lightgreen")

# Scatterplot 1: VOLUME vs RINGS
plot(mydata$RINGS, mydata$VOLUME,
     main = "VOLUME vs RINGS",
     xlab = "RINGS", ylab = "VOLUME",
     pch = 19, col = "blue")

# Scatterplot 2: WHOLE vs RINGS
plot(mydata$RINGS, mydata$WHOLE,
     main = "WHOLE vs RINGS",
     xlab = "RINGS", ylab = "WHOLE",
     pch = 19, col = "darkgreen")
```

**Essay Question (5 points) How well do you think these variables would perform as predictors of age?  Explain.**

***Answer: (
Based on the plot analysis-

1. Boxplots of VOLUME and WHOLE by CLASS
 - Here is a clear upward trend from Class A1 to A5 in both VOLUME and WHOLE.
 - The median values increase with CLASS, indicating that as abalones move to higher maturity classes, their size and weight increase.
 - The spread and presence of outliers increase with CLASS, showing growing variation among older individuals.
 - These patterns suggest a strong ordinal association between CLASS (which relates to age) and both variables.

2. Scatterplots of VOLUME and WHOLE vs RINGS
 - A positive correlation is observed for both VOLUME and WHOLE with RINGS.
 - The relationship appears nonlinear — growth accelerates in the middle age range (RINGS 6–15), then tapers or flattens slightly.
 - Higher variability is seen in older abalones (RINGS > 12), meaning these predictors may be less reliable at estimating older ages precisely.Both VOLUME and WHOLE weight increase as the CLASS changes from A1 to A5.

In conclusion, both VOLUME and WHOLE weight can be useful for predicting the age of abalones (measured by RINGS):
	- As abalones get older, their VOLUME and WHOLE weight generally increase, which means these are good indicators of age.
	- This is clearly seen in the boxplots, where older classes (A4, A5) have higher values.
	-	In the scatterplots, we see that as RINGS go up, VOLUME and WHOLE weight also go up — especially between RINGS 5 and 15.

However:
	-	The data becomes more spread out for older abalones. So, while these variables work well for younger and middle-aged abalones, they are less accurate for predicting the age of older ones.
	-	VOLUME might be slightly better than WHOLE, because it includes length, diameter, and height — giving a fuller picture of the abalone’s size.

Overall, VOLUME and WHOLE are good predictors of age, especially for younger abalones, but they become less reliable as abalones get older.

)***

-----

***### Section 5: (12 points) Getting insights regarding different groups in the data.***

(5)(a) (2 points) Use *aggregate()* with "mydata" to compute the mean values of VOLUME, SHUCK and RATIO for each combination of SEX and CLASS. Then, using *matrix()*, create matrices of the mean values. Using the "dimnames" argument within *matrix()* or the *rownames()* and *colnames()* functions on the matrices, label the rows by SEX and columns by CLASS. Present the three matrices (Kabacoff Section 5.6.2, p. 110-111). The *kable()* function is useful for this purpose.  You do not need to be concerned with the number of digits presented.

```{r Part_5a}
library(knitr)

# Compute means for each combination of SEX and CLASS
agg_data <- aggregate(cbind(VOLUME, SHUCK, RATIO) ~ SEX + CLASS, data = mydata, FUN = mean)

# Create matrices from the aggregated data
volume_matrix <- matrix(agg_data$VOLUME, nrow = length(unique(mydata$SEX)), byrow = TRUE,
                        dimnames = list(levels(mydata$SEX), levels(mydata$CLASS)))
shuck_matrix <- matrix(agg_data$SHUCK, nrow = length(unique(mydata$SEX)), byrow = TRUE,
                       dimnames = list(levels(mydata$SEX), levels(mydata$CLASS)))
ratio_matrix <- matrix(agg_data$RATIO, nrow = length(unique(mydata$SEX)), byrow = TRUE,
                       dimnames = list(levels(mydata$SEX), levels(mydata$CLASS)))

# Display the matrices using kable()
kable(volume_matrix, caption = "Mean VOLUME by SEX and CLASS")
kable(shuck_matrix, caption = "Mean SHUCK by SEX and CLASS")
kable(ratio_matrix, caption = "Mean RATIO by SEX and CLASS")
  
```

(5)(b) (3 points) Present three graphs. Each graph should include three lines, one for each sex. The first should show mean RATIO versus CLASS; the second, mean VOLUME versus CLASS; the third, mean SHUCK versus CLASS. This may be done with the 'base R' *interaction.plot()* function or with ggplot2 using *grid.arrange()*.

```{r Part_5b, fig.width = 9}
# Set up the plotting area: 3 plots vertically
par(mfrow = c(3, 1), mar = c(4, 4, 2, 1))

# Plot 1: Mean RATIO vs CLASS by SEX
interaction.plot(x.factor = mydata$CLASS, 
                 trace.factor = mydata$SEX,
                 response = mydata$RATIO,
                 fun = mean,
                 type = "b", col = c("blue", "green", "red"),
                 pch = c(1, 2, 3),
                 xlab = "CLASS", ylab = "Mean RATIO",
                 main = "Mean RATIO vs CLASS by SEX")

# Plot 2: Mean VOLUME vs CLASS by SEX
interaction.plot(x.factor = mydata$CLASS, 
                 trace.factor = mydata$SEX,
                 response = mydata$VOLUME,
                 fun = mean,
                 type = "b", col = c("blue", "green", "red"),
                 pch = c(1, 2, 3),
                 xlab = "CLASS", ylab = "Mean VOLUME",
                 main = "Mean VOLUME vs CLASS by SEX")

# Plot 3: Mean SHUCK vs CLASS by SEX
interaction.plot(x.factor = mydata$CLASS, 
                 trace.factor = mydata$SEX,
                 response = mydata$SHUCK,
                 fun = mean,
                 type = "b", col = c("blue", "green", "red"),
                 pch = c(1, 2, 3),
                 xlab = "CLASS", ylab = "Mean SHUCK",
                 main = "Mean SHUCK vs CLASS by SEX")

```

**Essay Question (2 points):  What questions do these plots raise?  Consider aging and sex differences.**

***Answer: (
These plots raise several important questions regarding aging patterns and sex differences in abalones:
	1. Why do Infant abalones (I) consistently have the lowest values of RATIO, VOLUME, and SHUCK across all CLASS levels?
	- This may indicate that the growth trajectory of Infants is fundamentally different, or that CLASS classification does not fully capture development stages for Infants.
	- Are Infants being misclassified into older CLASS groups?

  2.	Why do Female (F) abalones generally show higher mean VOLUME and SHUCK values compared to Males (M), especially in the older classes (A4 and A5)?
	-	This suggests possible sex-based growth differences, with Females potentially growing larger or having more mass relative to volume.
	-	Are these differences due to biological growth rates or reproductive adaptations?

	3. Why does RATIO decline for all sexes in higher CLASS levels, especially for Infants and Males?
	-	This could indicate a relative reduction in meat (SHUCK) to body size (VOLUME) as abalones age.
	-	Does aging lead to an increase in shell growth without a proportional increase in meat weight?

	4. Why do Males and Females converge in RATIO in younger classes but diverge in older ones?
	-	This pattern may reflect early similarities in development that diverge due to sexual dimorphism or energy allocation differences during aging.

	5. Is CLASS a reliable proxy for age for all sexes, especially Infants?
	-	Since Infants show unusually low values even in older CLASS groups, this questions whether CLASS categories fully align with biological age or maturity for this group.

These insights suggest the need for further biological investigation into growth dynamics, sex differentiation, and classification accuracy in the dataset.
)***

5(c) (3 points) Present four boxplots using *par(mfrow = c(2, 2)* or *grid.arrange()*. The first line should show VOLUME by RINGS for the infants and, separately, for the adult; factor levels "M" and "F," combined. The second line should show WHOLE by RINGS for the infants and, separately, for the adults. Since the data are sparse beyond 15 rings, limit the displays to less than 16 rings. One way to accomplish this is to generate a new data set using subset() to select RINGS < 16.  Use ylim = c(0, 1100) for VOLUME and ylim = c(0, 400) for WHOLE.  If you wish to reorder the displays for presentation purposes or use ggplot2 go ahead.

```{r Part_5c}
# Filter data for RINGS < 16 only
mydata_subset <- subset(mydata, RINGS < 16)

# Ensure SEX and CLASS are factors
mydata_subset$SEX <- as.factor(mydata_subset$SEX)
mydata_subset$CLASS <- as.factor(mydata_subset$CLASS)

# Set up 2x2 plotting layout
par(mfrow = c(2, 2), mar = c(4, 4, 3, 1))

# 1st Plot: VOLUME by RINGS for INFANT only
boxplot(VOLUME ~ RINGS, 
        data = subset(mydata_subset, SEX == "I"),
        main = "VOLUME vs RINGS (Infant)",
        xlab = "RINGS", 
        ylab = "VOLUME", 
        ylim = c(0, 1100), 
        col = "lightblue")

# 2nd Plot: VOLUME by RINGS for ADULTS (M + F)
boxplot(VOLUME ~ RINGS, 
        data = subset(mydata_subset, SEX %in% c("M", "F")),
        main = "VOLUME vs RINGS (Adults)",
        xlab = "RINGS", 
        ylab = "VOLUME", 
        ylim = c(0, 1100), 
        col = "salmon")

# 3rd Plot: WHOLE by RINGS for INFANT only
boxplot(WHOLE ~ RINGS, 
        data = subset(mydata_subset, SEX == "I"),
        main = "WHOLE vs RINGS (Infant)",
        xlab = "RINGS", 
        ylab = "WHOLE WEIGHT", 
        ylim = c(0, 400), 
        col = "lightgreen")

# 4th Plot: WHOLE by RINGS for ADULTS (M + F)
boxplot(WHOLE ~ RINGS, 
        data = subset(mydata_subset, SEX %in% c("M", "F")),
        main = "WHOLE vs RINGS (Adults)",
        xlab = "RINGS", 
        ylab = "WHOLE WEIGHT", 
        ylim = c(0, 400), 
        col = "orange")


```

**Essay Question (2 points):  What do these displays suggest about abalone growth?  Also, compare the infant and adult displays.  What differences stand out?**

***Answer: (
The boxplots provide meaningful insights into abalone growth patterns -	
  -	Volume and Whole weight both increase with RINGS (age), showing a strong positive relationship between physical size and age.
	-	Growth appears nonlinear:
	- Rapid increases occur between RINGS 3-10.
	-	Beyond RINGS 11, growth slows or plateaus, especially in VOLUME.

Comparing INFANTS vs ADULTS:

1. Growth magnitude:
	-	ADULTS (M + F) show much higher median and upper range in both VOLUME and WHOLE WEIGHT.
	-	INFANTS have lower and more tightly clustered values — indicating slower or early-stage growth.

2. Variability:
	-	Adults exhibit greater variability across RINGS, especially after age 8. Box heights are taller with more outliers.
	-	Infants have narrower boxes—their growth is more consistent and predictable early in life.

3. Outliers:
	-	Adults show more outliers in higher RINGS, possibly due to environmental or genetic variability in mature growth.
	-	Infants show fewer outliers, suggesting a more uniform growth pattern at younger ages.
	
)***

-----

***### Section 6: (11 points) Conclusions from the Exploratory Data Analysis (EDA).***

**Essay Question 1) (5 points)	Based solely on these data, what are plausible statistical reasons that explain the failure of the original study? Consider to what extent physical measurements may be used for age prediction.**

***Answer:(
There are several plausible statistical explanations for why the original study may have failed in accurately predicting abalone age from physical characteristics:

	- High variability within age groups:
  From the boxplots (Images 2, 8), both VOLUME and WHOLE measurements show increasing variability with age (RINGS). Especially in adults, the spread is wide even within the same age group, making precise prediction difficult.

	- Non-normality of RATIO:
  From the histograms, boxplots, and Q-Q plots by sex (Image 5), the RATIO variable deviates from normality with outliers and skewness. Since many models assume normality, this affects predictive reliability.

	- Overlap between groups:
  Scatterplots (Images 3, 4) show a strong correlation between WHOLE, VOLUME, and SHUCK weights, but there’s significant overlap across CLASS levels. This makes it hard to clearly distinguish between age classes using just physical        measures.

	- Nonlinear growth:
  Plots (Images 2, 8, 10) show that abalone growth in terms of WHOLE and VOLUME increases with RINGS, but the relationship plateaus for older individuals. That means prediction gets less accurate with age.

	- Sex-based variation:
  The interaction plots (Image 1) and summary matrices (Image 7–9) reveal that sex significantly influences growth patterns. Without accounting for sex differences, predictions may be systematically biased.

	- Misclassification or inconsistency in CLASS:
  The barplot of counts by SEX and CLASS (Image 1) shows an unexpected number of infants in higher classes. This suggests possible misclassification or inconsistency in how CLASS reflects age.

 Conclusion: While physical characteristics are correlated with age, overlap, variability, and biological differences limit their accuracy. This likely explains the failure of the original study’s predictive model.

 )***
 
**Essay Question 2) (3 points) Do not refer to the abalone data or study.  If you were presented with an overall histogram and summary statistics from a sample of some population or phenomenon and no other information, what questions might you ask before accepting them as representative of the sampled population or phenomenon?**

***Answer: (
Before accepting summary statistics and a histogram as representative of a population or phenomenon, I might have asked:

  1.How was the sample collected?
  - Was the sampling method random, stratified, or convenience-based? Sampling bias can lead to misleading conclusions.
  
  2.What is the sample size?
  - A small sample size may not capture the true variability or structure of the population.
  
  3.Does the histogram show skewness or outliers?
  - Averages and standard deviations can be misleading in the presence of skewed distributions or extreme values.
  
  4. Are there any known subgroups?
  - If the population includes multiple categories (e.g., age groups, regions), summary statistics for the overall sample might hide important patterns.

	5.Is there missing data or data quality issues?
  - Summary statistics may be based on incomplete or inconsistent records. Knowing data quality is essential before trusting the results.

	6.What is the context or purpose of analysis?
  - Without understanding the question being asked, it’s hard to judge whether the summary captures relevant information.
 )***

**Essay Question 3)  (3 points)	Do not refer to the abalone data or study.  What do you see as difficulties analyzing data derived from observational studies? Can causality be determined?  What might be learned from such studies?**

***Answer: (
Observational studies pose several challenges for analysis, particularly when it comes to establishing causality. Unlike experimental studies, observational data is collected without random assignment or controlled conditions. As a result, it is difficult if not impossible to draw definitive causal conclusions from such data. Instead, we can only identify associations or correlations between variables.

One major difficulty is the presence of confounding variables - unobserved factors that may influence both the independent and dependent variables. For example, a relationship between two variables might actually be driven by a third, unmeasured factor. Without the ability to control for all confounders, the risk of misleading interpretations increases.

Another concern is selection bias. Since participants in observational studies are not randomly selected or assigned, the sample may not represent the broader population. This limits the generalizability of findings. Measurement bias may also arise if the tools or methods used to collect data vary in accuracy or consistency across observations.

Despite these limitations, observational studies are incredibly valuable. They allow researchers to study real-world behavior and outcomes where experimentation is not feasible due to ethical, practical, or financial constraints. These studies can uncover meaningful patterns, inform policy decisions, and generate hypotheses for further experimental testing.

In summary, while observational studies do not support causal claims on their own, they play a crucial role in exploratory research and can provide strong evidence for relationships worth investigating in more controlled settings.
)***
